---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# PhaseTypeR: general-purpose phase-type functions

<!-- badges: start -->
[![DOI](https://joss.theoj.org/papers/10.21105/joss.05054/status.svg)](https://doi.org/10.21105/joss.05054)
[![R-CMD-check](https://github.com/rivasiker/PhaseTypeR/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/rivasiker/PhaseTypeR/actions/workflows/check-standard.yaml)
[![CRAN](https://www.r-pkg.org/badges/version/PhaseTypeR)](https://CRAN.R-project.org/package=PhaseTypeR)
[![devel Version](https://img.shields.io/badge/devel-1.0.5.9000-brightgreen.svg)](https://github.com/rivasiker/PhaseTypeR)
[![downloads](https://cranlogs.r-pkg.org/badges/grand-total/PhaseTypeR)](https://CRAN.R-project.org/package=PhaseTypeR)
[![Codecov test coverage](https://codecov.io/gh/rivasiker/PhaseTypeR/branch/master/graph/badge.svg)](https://app.codecov.io/gh/rivasiker/PhaseTypeR?branch=master)
<!-- badges: end -->


This package implements core functions from phase-type theory. Its general functions are useful for a wide-ranging variety of contexts. `PhaseTypeR` can be used to model continuous and discrete phase-type distributions, both univariate and multivariate. The package includes functions for outputting the mean and (co)variance of phase-type distributions; their density, probability and quantile functions; functions for random draws; functions for reward-transformation; and functions for plotting the distributions as networks. 

## Installation

You can install `PhaseTypeR` from CRAN via:

``` r
install.packages("PhaseTypeR")
```

If you install devtools in your R environment with `install.packages("devtools")`, the development version of the package can be installed with the following command:

``` r
devtools::install_github("rivasiker/PhaseTypeR")
```

These installations should handle the missing dependencies, but you can manually install them by running the following:

``` r
install.packages("expm")
install.packages("igraph")
```

## Basic example

This is a basic example for the univariate continuous phase-type distribution.

```{r warning=FALSE, message=FALSE}
# Load the package
library(PhaseTypeR)

# Define the sub-intensity rate matrix
subintensity_matrix <- matrix(c(-1.5,  0,  0,
                                 1.5, -1,  0,
                                  0,   1, -0.5), ncol = 3)
# Define the initial probabilities
initial_probabilities <- c(0.9, 0.1, 0)

# Create continuous phase-type object
ph <- PH(subintensity_matrix, initial_probabilities)
```

Summary of the object:

```{r}
summary(ph)
```


Density function:

```{r}
dPH(1:5, ph)
```

Cumulative density function:

```{r}
pPH(1:5, ph)
```

Quantile function:

```{r}
qPH(c(0.25, 0.5, 0.75), ph)
```

Random sampling:

```{r}
set.seed(0)
rPH(6, ph)
```

## Learn more

You can check out the full functionality of `PhaseTypeR` and its application to 
population genetics in the following guides:

  - [Getting Started](https://rivasiker.github.io/PhaseTypeR/articles/PhaseTypeR.html)
  - [Using PhaseTypeR for population genetics](https://rivasiker.github.io/PhaseTypeR/articles/a2_pop_gen_iker.html)
  - [Using PhaseTypeR for the site frequency spectrum](https://rivasiker.github.io/PhaseTypeR/articles/a3_phasetypeR_SFS.html)
  
## Report an issue

If you have found a bug, you can file an issue at <https://github.com/rivasiker/PhaseTypeR/issues>. Please, make sure to read the [contribution guidelines](.github/CONTRIBUTING.md) and the [code of conduct](.github/CODE_OF_CONDUCT.md) before making any changes. 
