---
title: "Phasty v0.1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phasty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, setup}
library(phasty)
```

# Informations

Please report any issue on the git page of the package ...

# Introduction

What is the topic
why a package 
based on what
What does it contains

# The phase-type class object

The package use several new classes for describing a phase-type distributed object, which are :"disc_phase_type", "cont_phase_type", "mult_disc_phase_type", "mult_cont_phase_type". Which can be obtain using the following functions respectively:

```{r, eval = F}
X1 <- disc_phase_type(subint_mat = NULL, init_probs = NULL)

X2 <- cont_phase_type(subint_mat = NULL, init_probs = NULL)

Y1 <- mult_disc_phase_type(subint_mat = NULL, init_probs = NULL, reward_mat = NULL)

Y2 <- mult_cont_phase_type(subint_mat = NULL, init_probs = NULL, reward_mat = NULL)
```

Those classes will be necessary later on to recognise the type of object for function like mean or var.

# The univariate distribution
## The continuous univariate phase-type
## The discrete univariate phase-type

# The multivariate distribution
## The continuous multivariate
## The discrete multivariate


# The discrete phase-type distribution


