---
title: "pop_gen_iker"
output: 
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{phasty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Introduction to the package

`phasty` is a flexible aand efficient package for using phase-type theory in R. This vignette describes how to use the general-purpose phase-type functions for modelling common statistics in population genomics at the finest level. The formulation of general phase-type theory can be consulted in Bladt and Nielsen's 'Matrix-Exponential Distributions in Applied Probability', and the notation for this vignette is also adopted from this book. On the other hand, the theory behind applying phase-type theory in population genomics is based on Hobolth et al (2018).

Do not hesitate to run `?phasty` for a quick summary of the available general-purpose functions, or to open the help files for the individual functions. 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = F,
  error = F
)
```

```{r setup}
library(phasty)
```


# 2. Continuous phase-type distributions: $T_{MRCA}$ and $T_{Total}$

## 2.1. Theoretical background

In an evolutionary tree the time until two sequences coalesce $T_i$ can be measured in number of generations $R_i$ divided by the population size $N$, this is, $T_i=R_i/N$. $T_i$ can easily be proven to approximate to an exponential distribution with rate $\binom{i}{2}$.

In order to understand the evolutionary history of sequences two additional quantities can be defined --namely the time until the most recent common ancestor $T_{MRCA}$ and the total tree length $T_{Total}$. $T_{MRCA}$ will simply be the sum of all times until two sequences coalesce, in other words $T_{MRCA}=T_n+T_{n-1}+...+T_2$, where $T_i\sim\text{exp}(\binom{i}{2})$. $T_{Total}$, on the other hand, takes into account the length of all possible branches, so $T_{Total}=nT_n+(n-1)T_{n-1}+...+2T_2$ and, thus, $iT_i\sim \text{exp}(\frac{i-1}{2})$.

The mean and variance of these two quantities can be derived relatively easily. Defining their distribution, however, is more challenging since both $T_{MRCA}$ and $T_{Total}$ are sums of independent exponentially distributed variables with different rates. Their distribution can be computed as a series of convolutions, but their formulation, application and interpretation might be challenging for the average population geneticist.

Instead, we can think of the sum of exponential distributions as a continuous-time Markov chain, where coalescent events are represented as Markov jumps with rate $T_i$ for $T_{MRCA}$ and $iT_i$ for $T_{Total}$. The Markov chain will end with an absorbing state, which in both cases will be the MRCA. 

The Markov chain can be represented using phase-type theory, where the jump rates are defined with a sub-intensity matrix $T$ and the initial distribution will be defined as a row vector $\pi$. If we define $\tau$ as the smallest time (or length) to reach the absorbing state, then $\tau\sim PH(\pi,T)$. This continuous phase-type distribution has well-documented and easy-to-implement formulas for the expectation, the variance, the survival function, the distribution function and the density function. Moreover, since both $\pi$ and $T$ can easily be speficied, we can represent evolutionary histories that do not follow the standard coalescent model and still use the same phase-type formulas. 

`phasty` contains an efficient implementation of continuous phase-type distributions. This section shows how to create phase-type representations of $T_{MRCA}$ and $T_{Total}$ under the Kingman's coalescent model, and it provides some guidelines for modeling these quantities for other non-standard coalescent models. 



## 2.2. Time until the most recent common ancestor ($T_{MRCA}$)

Following Hobolth et al. (2018), we can build a phase-type representation of $T_{MRCA}$ by defining the sub-intensity matrix, which each row and column will represent the states of the coalescent process. 

For example, for Kingman's coalescent the sub-intensity matrix can be defined as:


```{r}

subint_mat_t_mrca <- function(n) {
  # Define an matrix filled with zeros
  subint_mat <- matrix(c(0), nrow = n-1, ncol = n-1)
  for (i in 1:n-1) {
    # Save negative exponential rates in the diagonal
    subint_mat[i, i] <- - (n - i+1)*(n - i)/2
    if (i < n-1) {
      # Define the exponential jumping rates
      subint_mat[i, i+1] <- -subint_mat[i, i]
    }
  }
  subint_mat
}

```

For a sample size of $n=4$, the sub-intensity matrix can be generated using the function above, and the initial probabilities can be defined as a vector of length $n-1$, starting with one and the rest entries filled with zeros:

```{r}

n <- 4
subint_mat <- subint_mat_t_mrca(n)
init_probs <- matrix(c(1, rep(0,n-2)), 1, n-1)

```

Now we can use a `cont_phase_type` class of the `phasty` package to generate a phase-type representation of $T_{MRCA}$ by using `cont_phase_type()`:

```{r}

t_mrca_4 <- cont_phase_type(subint_mat, init_probs)

t_mrca_4

```

There are a number of methods associated with the class `cont_phase_type`. For example, the mean for $T_{MRCA}$ can be computed using: 

```{r}

mean(t_mrca_4)

```

While the variance can be calculated using:

```{r}

var(t_mrca_4)

```


## 2.3. Total tree lenght ($T_{Total}$)

In a similar way to $T_{MRCA}$, the total tree length or $T_{Total}$ can also be represented using phase-type theory (Hobolth et al. 2018).

For Kingman's coalescent the sub-intensity matrix can be defined as:

```{r}

subint_mat_t_total <- function(n) {
  subint_mat = matrix(c(0), nrow = n-1, ncol = n-1)
  for (i in 1:n-1) {
    subint_mat[i, i] = - 0.5 * (n - i)
    if (i < n-1) {
      subint_mat[i, i+1] = -subint_mat[i, i]
    }
  }
  subint_mat
}

```

For a sample size of $n=4$, the sub-intensity matrix and the initial probabilities can be defined as:

```{r}

n <- 4
subint_mat <- subint_mat_t_total(n)
init_probs <- matrix(c(1, rep(0,n-2)), 1, n-1)

```


$T_{Total}$ can also be represented using the `cont_phase_type` class:

```{r}

t_total_4 <- cont_phase_type(subint_mat, init_probs)

t_total_4

```

Without further ado, we can use the exact same phase-type formulas (and thus `phasty` functions) for calculating the mean and variance of $T_{Total}$:

```{r}

mean(t_total_4)
var(t_total_4)

```


## 2.4. Density, distribution and quantile functions 

`phasty` also includes the density function (`dphtype()`), quantile function (`qphtype()`), distribution function (`pphtype()`) and random draw generator (`rphtype()`) for continuous phase-type distributions. We can therefore apply these functions for our phase-type representation of $T_{MRCA}$ and $T_{Total}$. 

For example, the density function:

```{r}
dphtype(0.5, t_mrca_4)
dphtype(1:3, t_total_4)
```

```{r}
par(mfrow = c(1, 2))
x <- seq(0,10,0.1)
y <- dphtype(x, t_mrca_4)
plot(x, y, type = 'l')
lines(x, y)
title('Density function of T_MRCA')
x <- seq(0,20,0.1)
y <- dphtype(x, t_total_4)
plot(x, y, type = 'l')
lines(x, y)
title('Density function of T_total')
```

The quantile function:

```{r}
qphtype(0.5, t_mrca_4)
qphtype(c(0.25, 0.75), t_total_4)
```

```{r}
par(mfrow = c(1, 2))
x <- seq(0,0.99,0.01)
y <- qphtype(x, t_mrca_4)
plot(x, y, type = 'l')
lines(x, y)
title('Quantile function of T_MRCA')
x <- seq(0,0.99,0.01)
y <- qphtype(x, t_total_4)
plot(x, y, type = 'l')
lines(x, y)
title('Quantile function of T_total')
```

The probability function:

```{r}
pphtype(0.5, t_mrca_4)
pphtype(c(0.25, 0.75), t_total_4)
```

```{r}
par(mfrow = c(1, 2))
x <- seq(0,6,0.1)
y <- pphtype(x, t_mrca_4)
plot(x, y, type = 'l')
lines(x, y)
title('Probability function of T_MRCA')
x <- seq(0,12,0.1)
y <- pphtype(x, t_total_4)
plot(x, y, type = 'l')
lines(x, y)
title('Probability function of T_total')
```



```{r}
rphtype(3, t_mrca_4)
rphtype(10, t_total_4)
```

```{r}
par(mfrow = c(2, 1))
x <- rphtype(10000, t_mrca_4)
hist(x, main = '10,000 random draws of T_MRCA', breaks=20)
x <- rphtype(10000, t_total_4)
hist(x, main = '10,000 random draws of T_total', breaks=20)
```

# 3. Discrete Phase-Type Distributions: $S_{total}$, $\xi_i$ and $\xi_{k+}$

## 3.1. Theoretical background

Some quantities in population genomics are discrete, such as the total number of segregating sites $S_{total}$ or other statistics related to the site-frequency spectrum (singletons, doubletons, tail statistic, etc.). Representing the site-frequency spectrum, or SFS, can be challenging if the sample size is large, if the data does not follow Kingman's coalescence or if we are trying to estimate the mutation parameter $\theta$ using non-standard estimators. 

Luckily, all these quantities can be represented using discrete phase-type theory. Similar to the continuous phase-type distribution being a generalization of exponentially distributed variables, the discrete phase-type distribution can be seen as a generalization of various geometric distributions. Each state in the continuous phase-type distribution will represent a stochastic process, such that the time until absorption of a random variable ($\tau$) follows an absorbing Markov chain. In this case we say that $\tau+1\sim DPH(\pi, T)$, where $T$ is the sub-intensity matrix that gathers the transition rates and $\pi$ is the vector of initial probabilities.

The discrete phase-type representation of $S_{total}$ can be built from the continuous phase-type representation of the total branch length. Folllowing theorem 3.5 in Hobolth et. al (2019), the total branch length can be discretized if the mutation parameter $\theta$ is supplied. 

On the other hand, singletons, doubletons and related variables (referred to as i-tons or $\xi_i$ throughout this vignette), require a more elaborate formulation. The idea behind consists in first building a continuous phase-type representation of the coalescent process, where the sub-intensity matrix will contain the rates of transition between all the possible branch types of all possible genealogies. Among all these branches, only a certain number will give rise to i-tons. We can calculate which branches these are, together with the weights associated to each of them. Using this information, we can reward-transform the continuous phase-type representation of the coalescent process, so it only represents those branches that give rise to a certain i-ton. Afterwards, we can discretize this representation using theorem 3.5 in Hobolth et. al (2019) and the mutation parameter $\theta$. 

Bear in mind that it is possible for some i-tons to never be observed. This means that the initial probability vector might not sum up to 1, because the variable might directly jump to the absorbing state. The probability that this happens is the so-called defect.

Because we have calculated the rewards for each of the i-tons, we can easily compute the reward for the sum of various i-tons. As an example, we can calculate a phase-type representation of the tail statistic, which is defined as $S_{k+} = \sum_{i = k}^{n-1} \xi_i$, by reward-transforming the phase-type representation of the coalescent by the sum of the rewards of individual i-tons. This tail statistic, together with  normalized singletons, is useful to distinguish between the traditional Kingsman coalescent and a multiple merger coalescent (Koskela (2018)).

Note that $S_{total}$ can also be represented by reward-transforming the coalescent process with the sum of the rewards for all the i-tons. However, this representation of $S_{total}$ is more inefficient, so the one explained above is used in the package. 

Because phase-type theory has well-defined functions for discrete phase-type distributions, the expectation, the variance, the survival function, the distribution function and the density function of all the variables defined above can be computed efficiently. `phasty` contains an efficient implementation of continuous phase-type distribution, together with wrapper functions for representing i-tons, tail statistics and the number of segregating sites. 

## 3.2. Total number of segregating sites ($S_{total}$)

The total number of segregating sites can be represented using a `disc_phase_type` class, this is, by means of a discrete phase-type distribution. For example, when `n`=4:

```{r}
stotal_ph = segsites(4, theta = 2)
summary(stotal_ph)
```

The class `disc_phase_type` is a list containing:

* reward-transformed matrix (denoted as `P` in Hobolth (2019)).
* vector of initial probabilities ($\pi$).
* defect (probability of not entering any transient state prior to absorption).

The variance and the expectation of any discrete phase-type distribution can be computed with `var()` and `mean()` respectively:

```{r}
mean(stotal_ph)
var(stotal_ph)
```


## 3.3. i-tons ($\xi_i$)

The i-tons $\xi_i$ can also be represented using a discrete phase-type distribution. First, we will define the coalescent process as a continuous phase-type distribution:

```{r}
coalescent <- kingsman(6)
```

The `kingsman()` function returns a `mult_phase_type` object which is a continuous phase-type distribution with a reward matrix `RewardM`, for later reward-transformation and discretization. 

If we want to obtain a representation of the tripletons $\xi_3$ when `n`=6, we can use the coalescent in the `itons()` function:

```{r}
itons_ph = itons(coalescent, i = 3, theta = 2)
summary(itons_ph)
```

## 3.4. Tail-statistic ($\xi_{k+}$)

In `phasty` the tail statistic is implemented using the coalescent and the `tailstat()` function. For example, $\xi_{3+}$ when `n`=5:

```{r}
coalescent <- kingsman(5)
summary(tailstat(coalescent, k = 3, theta = 2))
```

## 3.5. User-defined sub-intensity matrix

As with the class `cont_phase_type`, using the `disc_phase_type()` generator the user can specify the sub-intensity matrix, with optional initial probabilities (see `?disc_phase_type` for further information).

## 3.6. Density, distribution and quantile functions

`phasty` also contains the density function (`dphtype()`), quantile function (`qphtype()`), distribution function (`pphtype()`) and random draw generator (`rphtype()`) for the discrete phase-type distribution. For example, for $S_{total}$ when $\theta=3$:

```{r}
S_tot_ph <- segsites(10, theta = 3)
summary(S_tot_ph)
```

```{r}
dphtype(3, S_tot_ph)
```

```{r}
x <- 1:30
y <- dphtype(x, S_tot_ph)
plot(x, y)
title('Density function of S_total with theta=3 and n=10')
```


```{r}
pphtype(15, S_tot_ph)
```

```{r}
x <- 1:30
y <- pphtype(x, S_tot_ph)
plot(x, y)
title('Distribution function of S_total with theta=3 and n=10')
```

```{r}
qphtype(0.5, S_tot_ph)
```

```{r}
x <- seq(0,0.99,0.01)
y <- qphtype(x, S_tot_ph)
plot(x, y, type = 'l')
lines(x, y)
title('Quantile function of S_total with theta=3 and n=10')
```

```{r}
rphtype(3, S_tot_ph)
```

```{r}
x <- rphtype(10000, S_tot_ph)
hist(x, main = 'Random draws of S_total with theta=3 and n=10', breaks=20)
```
